Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-27 article-models-tt.tex
fixing bugs of tmext
commutative dependence diagrams

\begin{lem}
The following compatibility rules for two consecutive term extensions is valid:
\begin{align*}
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jterm{\Gamma}{\subst{\tmext{A}{P}{x}{u}}{Q}}{v}
    }
  { \jtermeq
      {\Gamma}
      {\ctxext{{A}{P}}{Q}}
      {\tmext{A}{{P}{Q}}{x}{{u}{v}}}
      {\tmext{{A}{P}}{Q}{{x}{u}}{v}}
    }
  \\
& \inference
  { \jterm{\Gamma}{\ctxext{{A}{P}}{Q}}{w}
    }
  { \jtermeq
      {\Gamma}
      {A}
      {\cprojfst{A}{P}{\cprojfst{\ctxext{A}{P}}{Q}{w}}}
      {\cprojfst{A}{\ctxext{P}{Q}}{w}}
    }
  \\
& \inference
  { \jterm{\Gamma}{\ctxext{{A}{P}}{Q}}{w}
    }
  { \jtermeq
      {\Gamma}
      {\subst{\cprojfst{A}{\ctxext{P}{Q}}{w}}{P}}
      {\cprojsnd{A}{P}{\cprojfst{\ctxext{A}{P}}{Q}{w}}}
      {\cprojfst{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}
    }
  \\
& \inference
    { \jterm{\Gamma}{\ctxext{{A}{P}}{Q}}{w}
      }
    { \jtermeq
        {\Gamma}
        {\subst{\cprojfst{\ctxext{A}{P}}{Q}{w}}{Q}}
        {\cprojsnd{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}
        {\cprojsnd{\ctxext{A}{P}}{Q}{w}}
      }
\end{align*}
\end{lem}

@@

>>>>

\begin{lem}
The following compatibility rules for two consecutive term extensions is valid:
\begin{align*}
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jterm{\Gamma}{\subst{\tmext{A}{P}{x}{u}}{Q}}{v}
    }
  { \jtermeq
      {\Gamma}
      {\ctxext{{A}{P}}{Q}}
      {\tmext{A}{{P}{Q}}{x}{{\subst{x}{P}}{\subst{x}{Q}}{u}{v}}}
      {\tmext{{A}{P}}{Q}{{A}{P}{x}{u}}{v}}
    }
  \\
& \inference
  { \jterm{\Gamma}{\ctxext{{A}{P}}{Q}}{w}
    }
  { \jtermeq
      {\Gamma}
      {A}
      {\cprojfst{A}{P}{\cprojfst{\ctxext{A}{P}}{Q}{w}}}
      {\cprojfst{A}{\ctxext{P}{Q}}{w}}
    }
  \\
& \inference
  { \jterm{\Gamma}{\ctxext{{A}{P}}{Q}}{w}
    }
  { \jtermeq
      {\Gamma}
      {\subst{\cprojfst{A}{\ctxext{P}{Q}}{w}}{P}}
      {\cprojsnd{A}{P}{\cprojfst{\ctxext{A}{P}}{Q}{w}}}
      {\cprojfst{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}
    }
  \\
& \inference
    { \jterm{\Gamma}{\ctxext{{A}{P}}{Q}}{w}
      }
    { \jtermeq
        {\Gamma}
        {\subst{\cprojfst{\ctxext{A}{P}}{Q}{w}}{Q}}
        {\cprojsnd{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}
        {\cprojsnd{\ctxext{A}{P}}{Q}{w}}
      }
\end{align*}
\end{lem}

@@

\subsubsection{Commuting dependence diagrams}
Now that we have introduced the notions of morphisms, composition and identity
terms, we can develop a diagramatic style of of displaying type dependencies
combined with morphisms. We give an informal, metatheoretical definition of
such diagrams by indicating what the various components mean. The definition
is informal because we will only use such diagrams occasionally to provide a
graphical indication of the situation in which we're working.

\begin{defn}
A diagram is said to be a \emph{dependency diagram in context $\Gamma$}
if it is built up according to the following steps:
\begin{itemize}
\item An ordinary arrow 
\begin{equation*}
\begin{tikzcd}
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
between two families $A$ and $B$ of contexts over $\Gamma$ indicates that
$f$ is a morphism from $A$ to $B$ in context $\Gamma$, i.e.~that we have the
judgment $\jhom{\Gamma}{A}{B}{f}$.
\item The set of double-headed arrows must form a forest where the root of
each tree is a family of contexts over $\Gamma$. 
\item A sequence of double-headed 
arrows
\begin{equation*}
\begin{tikzcd}
P_{n+1} \ar[fib]{r} & \cdots \ar[fib]{r} & P_1 \ar[fib]{r} & A
\end{tikzcd}
\end{equation*}
indicates that $P_1$ is a family of contexts over $\ctxext{\Gamma}{A}$, that
$P_2$ is a family of contexts over $\ctxext{{\Gamma}{A}}{P_1}$, etcetera.
\item A ladder of double-headed arrows
\begin{equation*}
\begin{tikzcd}
P_{n+1} \ar{r}{F_{n+1}} \ar[fib]{d} & Q_{n+1} \ar[fib]{d}\\
\vdots \ar[fib]{d} & \vdots \ar[fib]{d}\\
P_1 \ar{r}{F_1} \ar[fib]{d} & Q_1 \ar[fib]{d}\\
A \ar{r}{f} & B
\end{tikzcd}
\end{equation*}
indicates that $F_1$ is a morphism from $P_1$ to $Q_1$ \emph{over} $f$,
i.e.~that the judgment $\jfhom{\Gamma}{A}{B}{f}{P_1}{Q_1}{F_1}$ holds, that
$F_2$ is a morphism from $P_2$ to $Q_2$ over
the morphism $\tmext{}{}{\ctxwk{P_1}{f}}{F_1}$ from $\ctxext{A}{P_1}$ to
$\ctxext{B}{Q_1}$, etcetera.
\end{itemize}
Such a diagram is said to be commutative if the subdiagram consisting of only
the normal headed arrows is commutative in the usual sense (using judgmental
equality). Note that the ladders are inherently commutative.
\end{defn}

\subsubsection{Another special case of projections}
In this subsubsection we investigate the special case of a projection which
appears as a morphism from $\ctxext{{A}{P}}{\ctxwk{P}{Q}}$ to $\ctxext{A}{Q}$
in context $\Gamma$, where we assume to have the families 
$\jfam{{\Gamma}{A}}{P}$ and $\jfam{{\Gamma}{A}}{Q}$.

Note that we have the judgmental
equalities
\begin{align*}
\ctxwk{\ctxext{{A}{P}}{\ctxwk{P}{\mfam{A}}}}{\ctxext{A}{\mfam{A}}}
& \jdeq 
  \ctxext
    {\ctxwk{\ctxext{{A}{P}}{\ctxwk{P}{\mfam{A}}}}{A}}
    {\ctxwk{\ctxext{{A}{P}}{\ctxwk{P}{\mfam{A}}}}{\mfam{A}}}
  \\
& \jdeq
  \ctxext
    {\ctxwk{{P}{\mfam{A}}}{{\ctxext{A}{P}}{A}}}
    {\ctxwk{\ctxext{{A}{P}}{\ctxwk{P}{\mfam{A}}}}{\mfam{A}}}
\end{align*}
Note that we have the term $\ctxwk{{P}{\mfam{A}}}{\cprojfstf{A}{P}}$ of the
family $\ctxwk{{P}{\mfam{A}}}{{\ctxext{A}{P}}{A}}$. Therefore, we need to
find a term of type $\subst{\ctxwk{{P}{\mfam{A}}}{\cprojfstf{A}{P}}}
{\ctxwk{\ctxext{{A}{P}}{\ctxwk{P}{\mfam{A}}}}{\mfam{A}}}$. Note that we have
the judgmental equalities:
\begin{align*}
\subst
  {\ctxwk{{P}{\mfam{A}}}{\cprojfstf{A}{P}}}
  {\ctxwk{\ctxext{{A}{P}}{\ctxwk{P}{\mfam{A}}}}{\mfam{A}}}
& \jdeq
  \subst
    {\ctxwk{{P}{\mfam{A}}}{\cprojfstf{A}{P}}}
    {\ctxwk{{P}{\mfam{A}}}{{\ctxext{A}{P}}{\mfam{A}}}}
  \\
& \jdeq
  \ctxwk
    { {P}{\mfam{A}}
      }
    { \subst
        {\cprojfstf{A}{P}}
        {\ctxwk{\ctxext{A}{P}}{\mfam{A}}}
      }
  \\
& \jdeq
  \ctxwk
    { {P}{\mfam{A}}
      }
    { {P}{\mfam{A}}
      }
  \\
& \jdeq
  \ctxwk{P}{{\mfam{A}}{\mfam{A}}}
\end{align*}
We find the term $\ctxwk{P}{\idtm{\mfam{A}}}$ here. Thus we can now define
$\bar{\typefont{pr}}$ by:
\begin{equation}\label{barproj}
\jhomdefn
  {\Gamma}
  {{{A}{P}}{\mfam{A}}}
  {{A}{\mfam{A}}}
  {\bar{\typefont{pr}}}
  {\tmext{}{}{\ctxwk{{P}{\mfam{A}}}{\cprojfstf{A}{P}}}{\ctxwk{P}{\idtm{\mfam{A}}}}}
\end{equation}

\begin{lem}
We have the judgmental equality
\begin{equation*}
\jfameq
  {{{{\Gamma}{A}}{P}}{\ctxwk{P}{\mfam{A}}}}
  {\jcomp{}{\bar{\typefont{pr}}}{Q}}
  {\ctxwk{P}{Q}}
\end{equation*}
for any family $Q$ of contexts over $\ctxext{{\Gamma}{A}}{\mfam{A}}$ 
\end{lem}

### 2014-03-27 article-models-ttmodels.tex
commutative dependence diagrams

We consider the following diagram helpful:
\begin{equation*}
\begin{tikzcd}
\jcomp{}{\bar{\typefont{pr}}}{{}{\tfext{A}}{\mfam{P}}}
  \rar{}
  \ar[fib]{d}
  &
  \jcomp{}{\tfext{A}}{\mfam{P}}
    \rar{}
    \ar[fib]{d}{}
  &
  \mfam{P}
    \ar[fib]{d}{}
  \\
\jcomp{}{\bar{\typefont{pr}}}{{}{\tfext{A}}{P}}
  \rar{}
  \ar[fib]{d}{}
  &
  \jcomp{}{\tfext{A}}{P}
  \rar{}
  \ar[fib]{d}{}
  &
  P
  \ar[fib]{d}{}\\
\ctxext{{A}{P}}{\mfam{A}}
  \ar{r}[swap]{\bar{\typefont{pr}}}
  &
  \ctxext{A}{\mfam{A}}
  \ar{r}[swap]{\tfext{A}}
  &
  A
\end{tikzcd}
\end{equation*}
Here $\bar{\typefont{pr}}\jdeq\tmext{}{}{\cprojfstf{A}{P}}{\idtm{\mfam{A}}}$,
the unmarked horizontal arrows indicate identity terms and the vertical arrows
merely indicate that what's above is a family depending on what's below.

>>>>

We consider the following dependency diagram helpful:
\begin{equation*}
\begin{tikzcd}
\jcomp{}{\bar{\typefont{pr}}}{{}{\tfext{A}}{\mfam{P}}}
  \ar[fib]{d}
  &
  \jcomp{}{\tfext{A}}{\mfam{P}}
    \ar[fib]{d}{}
  &
  \mfam{P}
    \ar[fib]{d}{}
  \\
\jcomp{}{\bar{\typefont{pr}}}{{}{\tfext{A}}{P}}
  \ar[fib]{d}{}
  &
  \jcomp{}{\tfext{A}}{P}
  \ar[fib]{d}{}
  &
  P
  \ar[fib]{d}{}\\
\ctxext{{A}{P}}{\ctxwk{P}{\mfam{A}}}
  \ar{r}[swap]{\bar{\typefont{pr}}}
  &
  \ctxext{A}{\mfam{A}}
  \ar{r}[swap]{\tfext{A}}
  &
  A
\end{tikzcd}
\end{equation*}
Here the morphism 
$\bar{\typefont{pr}}$ should be explained (skip ahead to 
\autoref{barproj} for the definition).
